Optical characterization of gaps in directly bonded Si compound optics using infrared spectroscopy

### 2015-09-21 AO_bondedSi.tex
use the two-column format.  near final version
futz with the column formatting, and prepare content for tarball archive
respond to referee- typos and extraneous line

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                         File: osa-revtex4-1.tex                             %
%                        Date: April 15, 2013                                 %
%                                                                             %
%                              BETA VERSION!                                  %
%                   JOSA A, JOSA B, Applied Optics, Optics Letters            %
%                                                                             %
%            This file requires the substyle file osajnl4-1.rtx,              %
%                   running under REVTeX 4.1 and LaTeX 2e                     %
%                                                                             %
%                   USE THE FOLLOWING REVTeX 4-1 OPTIONS:                     %
% \documentclass[osajnl,twocolumn,showpacs,superscriptaddress,10pt]{revtex4-1}%
%                    %% Use 11pt for Applied Optics                           %
%                                                                             %
%               (c) 2013 The Optical Society of America                       %
%                                                                             %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\documentclass[osajnl,twocolumn,showpacs,superscriptaddress,10pt]{revtex4-1} %% use 11pt for Applied Optics
\documentclass[osajnl,preprint,showpacs,superscriptaddress,12pt]{revtex4-1} %% use 12pt for preprint option
\usepackage{amsmath,amssymb,graphicx}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{hyperref}
%\usepackage{longtable}

@@

Verification that an optical bond meets its requirements and diagnosis of possible problems in the bonding technique require an accurate assessment of the geometry and lateral and axial dimensions of any defects.  Since silicon is opaque in the optical, the most straightforward technique for detecting defects in the bond is to image them in the infrared.  Fresnel losses will cause unbonded regions to appear darker than regions where the bond is perfect.  In cases where the axial extent of the gap is a half-wavelength or more, Newton rings appear in the image of the gap.  X-ray topography \cite{Mitani1990} exploits the change in optical density along lines of sight to allow precise high spatial resolution measurements of the spatial extent of gaps with axial dimensions as small as a few nanometers but not of the axial extent itself. In ultrasound microscopy, the density change at the gap interface produces reflections of sound waves and enables mapping the lateral gap dimensions but also produces only limited information about the axial dimension \cite{2000RScI...71.1869G}.

We describe here a new technique for measuring the axial extent of small gaps, founded upon the ability to measure transmission as a function of wavelength with high accuracy.   A recent generation of stock near-infrared spectrophotometers like the Cary5000 from Agilent offer $0.2\;\%$ precision and good repeatability.  We use the wavelength dependence of the transmission and a model of the gap as a closely-spaced Fabry-Perot etalon to measure axial gaps down to a few tens of nanometers in size.  In order to verify the applicability and accuracy of the technique, we have used microlithography to create small artificial gaps of known size in wafers that we then bond and measure.

For all types of gaps and bubbles, our technique needs to provide an accurate assessment of the transmission loss at the operating wavelength.  For finite-sized bubbles, measurement of the axial dimension helps us to assess the effects of high-temperature annealing \cite{Horn2009, Masteika2014} on removal of hydrogen bubbles.  A wavelength-dependent transmission variation can also indicate the presence of large numbers of gaps that are too small to resolve spatially, as one might expect if there are many very small contaminants present along the interface.  Our technique must be able to characterize gaps with losses of a few percent at the operating wavelength.  A $2\;\%$ loss at $1.25\;\mu$m corresponds to a gap of $>25\;$nm at normal incidence and the gap size for a loss of this magnitude scales linearly with wavelength.  We therefore need to be able to measure transmission as a function of wavelength to a fraction of a percent, including all systematic errors, for devices operating near $1\;\mu$m and to a somewhat looser standard for devices operating at longer wavelengths.

\section{Si$-$Si Fabry P\`{e}rot measurement theory}
\label{secTheory}

\subsection{The Si$-$Si interface gap is a low finesse Fabry-P\`{e}rot etalon}

@@

We can treat a small Si$-$Si interface gap between two almost-bonded silicon parts as a low-Finesse Fabry-P\`{e}rot etalon\cite{2007fuph.book.....S}.  The key property of an etalon is that it has a cavity enclosed between two smooth and parallel reflective surfaces where the requirements for smoothness and parallelism depend on the reflective finesse of the cavity.  For the narrow air gaps we consider here and for the modest reflectivity of a Si-vacuum interface, most gaps at Si$-$Si bonds will form effective etalons over most of their area.  The transmission through a Fabry-P\`{e}rot depends on the wavelength of light, the reflectivity of the etalon sidewalls, and the size of the gap.  We compute the reflectivity for an Si$-$vacuum interface, which is a modest function of wavelength through the refractive index of Silicon, $n(\lambda, T)$ \cite{2006SPIE.6273E..77F}, as shown in Figure \ref{figSiIndexFinesse}.  We drop the subscripts on the refractive index $n(\lambda, T) \rightarrow n$ and the coefficient of finesse $F(\lambda, T) \rightarrow F$ for clarity.  The computation is simply Fresnel's law at normal incidence:

@@

We define the coefficient of finesse\cite{2007fuph.book.....S}, $F$ (Equation \ref{Eq:coeffF}), in the customary way to encapsulate the Fabry-P\`{e}rot etalon's dependence on reflectivity.  The right panel of Figure \ref{figSiIndexFinesse} shows the coefficient of finesse as a function of wavelength.  In this work, we assume transmission is at normal incidence and the refractive index of the gap is 1.0.  Si absorbs negligibly longward of $\lambda$ = 1250 nm, which we verified by comparing the transmission of Si reference samples of different thicknesses (Figure \ref{figSiAbsorbfig}).  We measured the ratio of transmissions of a 0.5 mm thick double side polished Si wafer and a 3.0 mm Si puck.  Figure \ref{figSiAbsorbfig} verifies a key assumption that we make when comparing the transmission of scrap Si of different provenance and thickness: double side polished 3 mm thick Si pucks and 0.5 mm thick Si wafers have indistinguishable transmission at the $<0.2$\% level long-ward of $1250\;$nm.  The Si refractive index is indistinguishable from sample to sample for the wavelength ranges we care about- the non-absorbing wavelengths greater than about 1250 nm.  \cite{2006SPIE.6273E..77F} cites the wide variety of values for refractive index from the literature as evidence that batch-specific Si should be used as a reference for cases in which absolute accuracy greater than $\pm5\times10^{-3}$ is desired.  An absolute deviation of $\pm5\times10^{-3}$ in refractive index corresponds to a Fresnel transmission difference of about 0.2\%, which is comparable to our measurement uncertainty.  In the next subsection we work out the wavelength dependent transmission for the air etalon\footnote{A note on nomenclature- we use the term \emph{gap} for the physical structure, and \emph{etalon} for the model of that structure.}

@@

which has an average value of about 53\%.  Note that this transmission is above a na\"ive value of $T_{DSP}=(1-R^2)$, which does not take into account multiple incoherent reflections, and is therefore an underestimate.  The top line in the upper panel of Figure \ref{figAbsoluteTrans} shows the value derived from Equation \ref{eqnAbsDSPtrans} in the wavelength range $1200-2500\;$nm.  If the Si substrate thickness is less than the coherence length for the given spectral bandwidth, and if its surface roughness is much smaller than the wavelength, our assumptions break down and the multiple reflections would interfere coherently, and the outer Si cavities would behave as a Fabry-P\`erot etalon.  We do not treat this case here since our focus is on thick substrates with small gaps between the interfaces.

\begin{figure}[!htbp]
\includegraphics[width=0.40\columnwidth]{figs/etalon_trans.pdf}
\caption{\label{figAbsoluteTrans}The family of normal-incidence curves for transmission as a function of vacuum wavelength for various gap sizes in bonded Si optics.  \emph{Top panel}: Absolute transmission through a Si substrate with no gap (top denim-blue solid curve), and small gaps of axial extent 20 nm (green dashed curve), 35 nm (red dash-dotted curve), and 60 nm (purple dotted curve).  The absolute transmission through Si depends on wavelength since the refractive index depends minutely on wavelength.  The curves assume that multiple reflections within the Si substrates add incoherently as discussed in detail in Appendix \ref{sec:Append-IMRTMM}.  \emph{Middle panel}: The ``Gap Transmission'', defined as the Absolute transmission through a sample normalized by the absolute transmission through a gapless sample. The gap sizes and colors are the same as in the top panel.  \emph{Bottom panel}: ``Gap Transmission'' for larger gaps, spanning 0-3500 nm.  The 3500 nm sized gap (cyan solid line), demonstrates characteristic Fabry-P\`erot etalon fringes.}
\end{figure}

@@

We evaluated our metrology method experimentally by embedding gaps of known sizes in the interface between two bonded Si substrates and measuring the effects of these gaps on IR transmission.  The substrate thicknesses ranged from 0.5 to 3.3 mm.  We denote substrates with thicknesses larger than 1 mm as ``pucks'' and substrates with thickness less than 1.0 mm thick as ``wafers''.  We distinguish the two classes of substrate thickness based on the anticipation that thin substrates will conform more easily to their bonding partner substrates.  There is also evidence that bond front propagation is slower in thick pucks \cite{2007ApOpt..46.6793H} than in wafers.  Our ultimate goal is to bond optics with thicknesses as large as 30 mm.

@@

The spectral sampling ranged from 2.0 to 10.0 nm.  We set the spectrometer slit width to produce a spectral resolution of 5.0 nm.  We expect smoothly varying spectral features for all conceivable interface gaps in Si bonded wafers.  Specifically, Fabry-P\`erot fringes attributable to interfacial gaps in bonded Si will be well-sampled with 5 nm spectral resolution, so long as the interface gap is less than $\sim$100 $\mu$m.  This maximum gap limitation brings our technique to the point where it overlaps in capability with other techniques since the axial extents of such large gaps would be easily measurable by other means, like IR imaging and fringe-counting.

@@

Figure \ref{figVG03full} shows a parameter determination for the measurement of a bond over a deep ($4000 \pm 200\;$nm) 100\% fill factor gap in part VG03.  The analysis yields a best fit consistent with 100\% fill factor, and a depth of $3960 \pm 2\;$nm, well within the uncertainty of the physical measurement (Table \ref{tbl_experiments}).  The spectrum in this region shows characteristic Fabry-P\`erot fringes of constructive and destructive interference.  The only tunable parameters of the model are the gap axial extent, $d$ and fill factor $f$.  No scaling has been applied to the curves in Figure \ref{figVG03full}.  Figure \ref{figVG03part} shows a fit for a region of the same bonded pair where a nominally $4000\;$nm deep, 2 mm wide trench crosses the field (about 1 mm $\times$ 10 mm) scanned by the Cary 5000.  Analysis of this transmission curve gave a best fit depth of $4094 \pm 4\;$nm and a pattern fill factor of $4.6 \pm 0.1\;\%$, both consistent with the part geometry.  These two results show that the formalism performs as expected on resonant gaps several wavelengths deep.

@@

The matrix for the silicon air gap is calculated in the following way.  We treat the air gap as a Fabry-P\`erot etalon.  We do not need to consider the microscopic coherent interactions with the etalon transmission, all of that information is encapsulated in these equations for a Fabry-P\`erot etalon model for the gap:

@@

with $\lambda$ the vacuum wavelength, $R$ the Fresnel reflection of silicon, and $F$ the coefficient of finesse.  The coefficient of finesse $F$ and the phase $\delta$ are the two parameters of the Fabry-P\`erot etalon.  The coefficient of Finesse encapsulates the Fresnel reflection and depends only on the Si refractive index which has only a small wavelength (and temperature) dependence.  The phase depends (Equation \ref{eq:phase}) on the wavelength $\lambda$ and $d$ the air gap spacing.  We assume the gap is lossless, i.e. $T_g+R_g=1$.  The incoherent scattering and transfer matrices for the gap are then:

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                         File: osa-revtex4-1.tex                             %
%                        Date: April 15, 2013                                 %
%                                                                             %
%                              BETA VERSION!                                  %
%                   JOSA A, JOSA B, Applied Optics, Optics Letters            %
%                                                                             %
%            This file requires the substyle file osajnl4-1.rtx,              %
%                   running under REVTeX 4.1 and LaTeX 2e                     %
%                                                                             %
%                   USE THE FOLLOWING REVTeX 4-1 OPTIONS:                     %
% \documentclass[osajnl,twocolumn,showpacs,superscriptaddress,10pt]{revtex4-1}%
%                    %% Use 11pt for Applied Optics                           %
%                                                                             %
%               (c) 2013 The Optical Society of America                       %
%                                                                             %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\documentclass[osajnl,twocolumn,showpacs,superscriptaddress,10pt]{revtex4-1} %% use 11pt for Applied Optics
\documentclass[osajnl,preprint,showpacs,superscriptaddress,12pt]{revtex4-1} %% use 12pt for preprint option
\usepackage{amsmath,amssymb,graphicx}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{hyperref}
%\usepackage{longtable}

@@

Verification that an optical bond meets its requirements and diagnosis of possible problems in the bonding technique require an accurate assessment of the geometry and lateral and axial dimensions of any defects.  Since silicon is opaque in the optical range, the most straightforward technique for detecting defects in the bond is to image them in the infrared.  Fresnel losses will cause unbonded regions to appear darker than regions where the bond is perfect.  In cases where the axial extent of the gap is a half-wavelength or more, Newton rings appear in the image of the gap.  X-ray topography \cite{Mitani1990} exploits the change in optical density along lines of sight to allow precise high spatial resolution measurements of the spatial extent of gaps with axial dimensions as small as a few nanometers but not of the axial extent itself. In ultrasound microscopy, the density change at the gap interface produces reflections of sound waves and enables mapping the lateral gap dimensions but also produces only limited information about the axial dimension \cite{2000RScI...71.1869G}.

We describe here a new technique for measuring the axial extent of small gaps, founded upon the ability to measure transmission as a function of wavelength with high accuracy.   A recent generation of stock near-infrared spectrophotometers like the Cary5000 from Agilent offer $0.2\;\%$ precision and good repeatability.  We use the wavelength dependence of the transmission and a model of the gap as a closely-spaced Fabry-P\'{e}rot etalon to measure axial gaps down to a few tens of nanometers in size.  In order to verify the applicability and accuracy of the technique, we have used microlithography to create small artificial gaps of known size in wafers that we then bond and measure.

For all types of gaps and bubbles, our technique needs to provide an accurate assessment of the transmission loss at the operating wavelength.  For finite-sized bubbles, measurement of the axial dimension helps us to assess the effects of high-temperature annealing \cite{Horn2009, Masteika2014} on removal of hydrogen bubbles.  A wavelength-dependent transmission variation can also indicate the presence of large numbers of gaps that are too small to be resolved spatially, as one might expect if there are many very small contaminants present along the interface.  Our technique must be able to characterize gaps with losses of a few percent at the operating wavelength.  A $2\;\%$ loss at $1.25\;\mu$m corresponds to a gap of $>25\;$nm at normal incidence and the gap size for a loss of this magnitude scales linearly with wavelength.  We therefore need to be able to measure transmission as a function of wavelength to a fraction of a percent, including all systematic errors, for devices operating near $1\;\mu$m and to a somewhat looser standard for devices operating at longer wavelengths.

\section{Si$-$Si Fabry P\'{e}rot measurement theory}
\label{secTheory}

\subsection{The Si$-$Si interface gap is a low finesse Fabry-P\'{e}rot etalon}

@@

We can treat a small Si$-$Si interface gap between two almost-bonded silicon parts as a low-Finesse Fabry-P\'{e}rot etalon\cite{2007fuph.book.....S}.  The key property of an etalon is that it has a cavity enclosed between two smooth and parallel reflective surfaces where the requirements for smoothness and parallelism depend on the reflective finesse of the cavity.  For the narrow air gaps we consider here and for the modest reflectivity of a Si-vacuum interface, most gaps at Si$-$Si bonds will form effective etalons over most of their area.  The transmission through a Fabry-P\'{e}rot depends on the wavelength of light, the reflectivity of the etalon sidewalls, and the size of the gap.  We compute the reflectivity for an Si$-$vacuum interface, which is a modest function of wavelength through the refractive index of Silicon, $n(\lambda, T)$ \cite{2006SPIE.6273E..77F}, as shown in Figure \ref{figSiIndexFinesse}.  We drop the subscripts on the refractive index $n(\lambda, T) \rightarrow n$ and the coefficient of finesse $F(\lambda, T) \rightarrow F$ for clarity.  The computation is simply Fresnel's law at normal incidence:

@@

We define the coefficient of finesse\cite{2007fuph.book.....S}, $F$ (Equation \ref{Eq:coeffF}), in the customary way to encapsulate the Fabry-P\'{e}rot etalon's dependence on reflectivity.  The right panel of Figure \ref{figSiIndexFinesse} shows the coefficient of finesse as a function of wavelength.  In this work, we assume transmission is at normal incidence and the refractive index of the gap is 1.0.  Si absorbs negligibly longward of $\lambda$ = 1250 nm, which we verified by comparing the transmission of Si reference samples of different thicknesses (Figure \ref{figSiAbsorbfig}).  We measured the ratio of transmissions of a 0.5 mm thick double side polished Si wafer and a 3.0 mm Si puck.  Figure \ref{figSiAbsorbfig} verifies a key assumption that we make when comparing the transmission of scrap Si of different provenance and thickness: double side polished 3 mm thick Si pucks and 0.5 mm thick Si wafers have indistinguishable transmission at the $<0.2$\% level long-ward of $1250\;$nm.  The Si refractive index is indistinguishable from sample to sample for the wavelength ranges we care about- the non-absorbing wavelengths greater than about 1250 nm.  \cite{2006SPIE.6273E..77F} cites the wide variety of values for refractive index from the literature as evidence that batch-specific Si should be used as a reference for cases in which absolute accuracy greater than $\pm5\times10^{-3}$ is desired.  An absolute deviation of $\pm5\times10^{-3}$ in refractive index corresponds to a Fresnel transmission difference of about 0.2\%, which is comparable to our measurement uncertainty.  In the next subsection we work out the wavelength dependent transmission for the air etalon\footnote{A note on nomenclature- we use the term \emph{gap} for the physical structure, and \emph{etalon} for the model of that structure.}

@@

which has an average value of about 53\%.  Note that this transmission is above a na\"ive value of $T_{DSP}=(1-R^2)$, which does not take into account multiple incoherent reflections, and is therefore an underestimate.  The top line in the upper panel of Figure \ref{figAbsoluteTrans} shows the value derived from Equation \ref{eqnAbsDSPtrans} in the wavelength range $1200-2500\;$nm.  If the Si substrate thickness is less than the coherence length for the given spectral bandwidth, and if its surface roughness is much smaller than the wavelength, our assumptions break down and the multiple reflections would interfere coherently, and the outer Si cavities would behave as a Fabry-P\'{e}rot etalon.  We do not treat this case here since our focus is on thick substrates with small gaps between the interfaces.

\begin{figure}[!htbp]
\includegraphics[width=0.40\columnwidth]{figs/etalon_trans.pdf}
\caption{\label{figAbsoluteTrans}The family of normal-incidence curves for transmission as a function of vacuum wavelength for various gap sizes in bonded Si optics.  \emph{Top panel}: Absolute transmission through a Si substrate with no gap (top denim-blue solid curve), and small gaps of axial extent 20 nm (green dashed curve), 35 nm (red dash-dotted curve), and 60 nm (purple dotted curve).  The absolute transmission through Si depends on wavelength since the refractive index depends minutely on wavelength.  The curves assume that multiple reflections within the Si substrates add incoherently as discussed in detail in Appendix \ref{sec:Append-IMRTMM}.  \emph{Middle panel}: The ``Gap Transmission'', defined as the Absolute transmission through a sample normalized by the absolute transmission through a gapless sample. The gap sizes and colors are the same as in the top panel.  \emph{Bottom panel}: ``Gap Transmission'' for larger gaps, spanning 0-3500 nm.  The 3500 nm sized gap (cyan solid line), demonstrates characteristic Fabry-P\'{e}rot etalon fringes.}
\end{figure}

@@

We evaluated our metrology method experimentally by embedding gaps of known sizes in the interface between two bonded Si substrates and measuring the effects of these gaps on IR transmission.  The substrate thicknesses ranged from 0.5 to 3.3 mm.  We denote substrates with thicknesses larger than 1 mm as ``pucks'' and substrates with thickness less than 1.0 mm thick as ``wafers''.  We distinguish the two classes of substrate thickness based on the anticipation that thin substrates will conform more easily to their bonding partner substrates.  There is also evidence that bond front propagation is slower in thick pucks \cite{2007ApOpt..46.6793H} than in wafers.

@@

The spectral sampling ranged from 2.0 to 10.0 nm.  We set the spectrometer slit width to produce a spectral resolution of 5.0 nm.  We expect smoothly varying spectral features for all conceivable interface gaps in Si bonded wafers.  Specifically, Fabry-P\'{e}rot fringes attributable to interfacial gaps in bonded Si will be well-sampled with 5 nm spectral resolution, so long as the interface gap is less than $\sim$100 $\mu$m.  This maximum gap limitation brings our technique to the point where it overlaps in capability with other techniques since the axial extents of such large gaps would be easily measurable by other means, like IR imaging and fringe-counting.

@@

Figure \ref{figVG03full} shows a parameter determination for the measurement of a bond over a deep ($4000 \pm 200\;$nm) 100\% fill factor gap in part VG03.  The analysis yields a best fit consistent with 100\% fill factor, and a depth of $3960 \pm 2\;$nm, well within the uncertainty of the physical measurement (Table \ref{tbl_experiments}).  The spectrum in this region shows characteristic Fabry-P\'{e}rot fringes of constructive and destructive interference.  The only tunable parameters of the model are the gap axial extent, $d$ and fill factor $f$.  No scaling has been applied to the curves in Figure \ref{figVG03full}.  Figure \ref{figVG03part} shows a fit for a region of the same bonded pair where a nominally $4000\;$nm deep, 2 mm wide trench crosses the field (about 1 mm $\times$ 10 mm) scanned by the Cary 5000.  Analysis of this transmission curve gave a best fit depth of $4094 \pm 4\;$nm and a pattern fill factor of $4.6 \pm 0.1\;\%$, both consistent with the part geometry.  These two results show that the formalism performs as expected on resonant gaps several wavelengths deep.

@@

The matrix for the silicon air gap is calculated in the following way.  We treat the air gap as a Fabry-P\'{e}rot etalon.  We do not need to consider the microscopic coherent interactions with the etalon transmission, all of that information is encapsulated in these equations for a Fabry-P\'{e}rot etalon model for the gap:

@@

with $\lambda$ the vacuum wavelength, $R$ the Fresnel reflection of silicon, and $F$ the coefficient of finesse.  The coefficient of finesse $F$ and the phase $\delta$ are the two parameters of the Fabry-P\'{e}rot etalon.  The coefficient of Finesse encapsulates the Fresnel reflection and depends only on the Si refractive index which has only a small wavelength (and temperature) dependence.  The phase depends (Equation \ref{eq:phase}) on the wavelength $\lambda$ and $d$ the air gap spacing.  We assume the gap is lossless, i.e. $T_g+R_g=1$.  The incoherent scattering and transfer matrices for the gap are then:
